---
title: "DATA607_Assignment5"
author: "Pei-Ming Chen"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
  pdf_document: default
---
Create a .CSV file that includes all of the information above. It is encouraged to use a “wide” structure similar to how the information appears above, so that I  can practice tidying and transformations as described below. Read the information from your.CSV file in to R,and use tidyr and dplyr as needed to tidy and transform the data.Perform analysis to compare the arrival delays for the two airlines.The code should be in an R Markdown file, posted to rpubs.com, and should include narrative descriptions of the data cleanup work, analysis, and conclusions.

### Packages download

```{r load-packages, message=FALSE}
library(tidyr)
library(dplyr)
```

### FLights data CSV. file was uploaded to my Github account. 

```{r}
flights <- "https://raw.githubusercontent.com/PMCformosa/Assignment-5/main/Data_Flights.csv"

flights_Data <- read.table(flights,header = TRUE, sep=",", na.strings = c("","NA"))

flights_Data 
```

### Tidy up the data 

```{r }
flights_Data[2,1] <- flights_Data[1,1]
flights_Data[3,] <- flights_Data[4,] 
flights_Data[5,1] <- flights_Data[4,1]
flights_Data[4,] <- flights_Data[5,] 
flights_Data <- flights_Data[-5,]

flights_Data 
```

### Rename the first and second column head
```{r}
colnames (flights_Data) [1] <- "Airline"
colnames (flights_Data) [2] <- "Arrival_status"

flights_Data

```

### Convert wide data format into a long one by using gather function of tidyr package
```{r}
flights_LongData <- gather(flights_Data, "city", "n", 3:7)

flights_LongData
```

```{r}
dplyr::glimpse(flights_LongData)
```
### Spread the elements of the Arrival_status column into two separate columns names "delayed" and "on time" by using the spread() function of the dplyr package
```{r}
Airline_data2 <- flights_LongData %>% spread(Arrival_status, n)
Airline_data2
```
### Rename the fourth column 
```{r}
colnames (Airline_data2)[4] <- "on_time"
Airline_data2
```
### Use the pipe operator to obtain mean and median values of delayed or on_time numbers
The result below showed that the mean number of delayed flights are 128.8 with a median number 83.5. And the mean number of on_time flights are 971.2 with a median 440. 
```{r}
Airline_data2 %>% summarise(mean = mean(delayed), median = median(delayed), n = n())
Airline_data2 %>% summarise(mean = mean(on_time), median = median(on_time), n = n())

```


### Use the pipe operator to obtain the rate of on_time flights 
The highest on_time rate is observed in ALASKA airline flying Phoenix. And the lowest on_time rate happened in AMWEST airline flying San.Francisco. 

```{r}
Airline_data2 <- mutate(Airline_data2, rate_on_time = on_time/(on_time+delayed))
Airline_data3 <- mutate(Airline_data2, rate_delayed = delayed/(on_time+delayed))
Airline_data3
```

```{r}
Airline_data2 %>% 
   group_by(Airline) %>% 
        dplyr::summarise(max = max(delayed), min=min(delayed),
        mean=mean(delayed), median=median(delayed))
```

### Conclusion 
From the data cleaning and analysis above, we can see that AMWEST airline had higher mean and median delayed flight numbers, compared to the numbers of ALASKA. AMWEST airline flying San.Francisco also had the lowest on_time rate.  
